---
title: "Sponge assemblage: biodiversity linear mixed effects models"
author: "Alberto Rovellini"
date: "22 November 2018"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
---

```{r, echo = FALSE, include=FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, width = 150)
```

```{r, message = FALSE, warning = FALSE}

library(abind)
library(plyr)
library(ggplot2)
library(reshape2)
library(BiodiversityR)
library(ggvegan)
library(dplyr)
library(nlme)
library(effects)

```

Read-in data.

```{r}

dataAllYears <- read.csv("//Staff/Home/SCIFAC/rovellal/DocumentsRedir//Projects/Chapter1/analysis/dataAllYears.csv")

names(dataAllYears) <- substr(names(dataAllYears), 2, nchar(names(dataAllYears)))

```

Set up functions to calculate evenness, richness and diversity of the assemblage.

```{r}

# let's find out what the evenness is doing here, and then model it against the abundance

# species richness here, or number of species

richness <- function(x) {
  R <- sum(x != 0)
  return(R)
}

# Shannon's index

shannons.diversity <- function(x) {
  S <- sum(x != 0) # numer of species in the quadrat, only non-zeroes (required! we don't deal with "hypothetical" species!)
  total.inds <- sum(x) # total sponge number
  non.zero <- x[x != 0]
  rel.abs <- non.zero/total.inds # relative abundances for each species
  H <- -sum(rel.abs*log(rel.abs)) # Shannon-Wiener's index
  return(H)
}

simpsons.evenness <- function(x) {
  S <- sum(x != 0) # numer of species in the quadrat, only non-zeroes (required! we don't deal with "hypothetical" species!)
  total.inds <- sum(x) # total sponge number
  rel.abs <- x/total.inds # relative abundances for each species
  D <- sum(rel.abs^2) # simpson's diversity
  E <- 1/(D*S) # simpson's evenness
  return(E)
}

# apply to data

richness.vector <- apply(dataAllYears, 2, richness)
diversity.vector <- apply(dataAllYears, 2, shannons.diversity)
evenness.vector <- apply(dataAllYears, 2, simpsons.evenness)

# melt for plotting

melt.evenness <- melt(evenness.vector, value.name = "Evenness")
index <- rownames(melt.evenness)
melt.evenness$Year <- as.numeric(substr(index, 1, 2))
melt.evenness$Site <- substr(index, 3, 4)
melt.evenness$Quadrat <- substr(index, nchar(index), nchar(index))  
melt.evenness$QuadratSite <- substr(index, nchar(index)-1, nchar(index))
melt.evenness$Richness <- melt(richness.vector, value.name = "Richness")[,1]
melt.evenness$Diversity <- melt(diversity.vector, value.name = "Diversity")[,1]

# bring in the total data

melt.evenness$Sponges <- melt(colSums(dataAllYears), value.name = "Sponges")[,1]

# adjust class of columns

melt.evenness$Site <- factor(melt.evenness$Site)
melt.evenness$Quadrat <- factor(melt.evenness$Quadrat )
melt.evenness$QuadratSite <- factor(melt.evenness$QuadratSite)

glimpse(melt.evenness)

```

## Mean values of biodiversity metrics

```{r}

biodiversity <- melt.evenness %>% group_by(
  Year
) %>% dplyr::summarize(
  Rich    = mean(Richness),
  Even    = mean(Evenness),
  Div     = mean(Diversity),
  Num     = mean(Sponges),
  Rich.SE = sd(Richness)/sqrt(length(Richness)),
  Even.SE = sd(Evenness)/sqrt(length(Evenness)),
  Div.SE  = sd (Diversity)/sqrt(length(Diversity)),
  Num.SE  = sd(Sponges)/sqrt(length(Sponges))
)

# make 2 one for means and one for errors (needed for melting later)

means <- biodiversity %>% dplyr::select(
  Year:Num
)

errors <- biodiversity %>% dplyr::select(
  Year, Rich.SE:Num.SE
)

# melt them

mean.div.plot <- melt(means, id.vars = "Year")
se.div.plot <- melt(errors, id.vars = "Year")

# bind them

plot.biodiversity <- cbind(mean.div.plot, se.div.plot[,ncol(se.div.plot)])

names(plot.biodiversity) <- c("year", "variable", "value", "SE") 

```

```{r}

mp <- ggplot(data = plot.biodiversity, aes(x = year, y = value, group = variable))+
  geom_point()+
  geom_line()+
  geom_errorbar(ymin = plot.biodiversity$value - plot.biodiversity$SE,
                ymax = plot.biodiversity$value + plot.biodiversity$SE)+
  facet_wrap(~ variable, scales = "free_y")
mp

```



## Linear models for biodiversity metrics. Using LMM here and quadrats as random factor.

### Richness

```{r, fig.width = 11, fig.height = 3.5}

null.rich <- nlme::lme(Richness ~ 1, 
                       random = ~ 1 | QuadratSite, 
                       data = melt.evenness, 
                       correlation = corCAR1(form = ~ Year | QuadratSite), 
                       method = "ML") # intercept model

sponge.rich <- update(null.rich, . ~ . + Sponges) # only sponges
site.rich <- update(null.rich, . ~ . + Site)
add.rich <- update(sponge.rich, . ~ . + Site)
all.rich <- update(add.rich, . ~ . + Sponges * Site)

# Backwards selection

anova(all.rich, add.rich) # interaction significant, P = 2e-04
anova(add.rich, site.rich) # sponges alone significant, P <.0001
anova(add.rich, sponge.rich) # site alone significant, P = 0.003

# fit variance structure

all.rich.var <- update(all.rich, weights = varIdent(form = ~ 1 | QuadratSite))

# re-fit final model with REML

rich.model <- update(all.rich.var, method = "REML")

summary(rich.model)

# Phi = 0.4403058

jpeg("//Staff/Home/SCIFAC/rovellal/DocumentsRedir/Projects/Defense/residuals/Ch3_LMM_S_Site_Year_SiteYear.jpeg", width = 11, height = 3.5, units = "in", res = 300) 
par(mfrow = c(1, 3))
plot(resid(rich.model, type = "normalized"))
hist(residuals(rich.model, type = "normalized"))
qqnorm(residuals(rich.model, type = "normalized"))
qqline(residuals(rich.model, type = "normalized")) # might be slightly skewed
dev.off()


```

### Diversity

```{r, fig.width = 12, fig.height = 5}

null.div <- nlme::lme(Diversity ~ 1, 
                       random = ~ 1 | QuadratSite, 
                       data = melt.evenness, 
                       correlation = corCAR1(form = ~ Year | QuadratSite), 
                       method = "ML") # intercept model

sponge.div <- update(null.div, . ~ . + Sponges) # only sponges
site.div <- update(null.div, . ~ . + Site)
add.div <- update(sponge.div, . ~ . + Site)
all.div <- update(add.div, . ~ . + Sponges * Site)

# Backwards selection

anova(all.div, add.div) # interaction significant, P = 0.05
anova(add.div, site.div) # sponges alone not significant, P = 0.9866
anova(add.div, sponge.div) # site alone significant, P = 0.0019

# fit variance structure

all.div.var <- update(all.div, weights = varIdent(form = ~ 1 | QuadratSite))

# re-fit final model with REML

div.model <- update(all.div.var, method = "REML")

summary(div.model)

# Phi = 0.3541138

jpeg("//Staff/Home/SCIFAC/rovellal/DocumentsRedir/Projects/Defense/residuals/Ch3_LMM_H_Site_Year_SiteYear.jpeg", width = 11, height = 3.5, units = "in", res = 300) 
par(mfrow = c(1, 3))
plot(resid(div.model, type = "normalized"))
hist(residuals(div.model, type = "normalized"))
qqnorm(residuals(div.model, type = "normalized"))
qqline(residuals(div.model, type = "normalized"))
dev.off()

```

### Evenness

```{r, fig.width = 12, fig.height = 5}

null.even <- nlme::lme(Evenness ~ 1, 
                       random = ~ 1 | QuadratSite, 
                       data = melt.evenness, 
                       correlation = corCAR1(form = ~ Year | QuadratSite), 
                       method = "ML") # intercept model

sponge.even <- update(null.even, . ~ . + Sponges) # only sponges
site.even <- update(null.even, . ~ . + Site)
add.even <- update(sponge.even, . ~ . + Site)
all.even <- update(add.even, . ~ . + Sponges * Site)

# Backwards selection

anova(all.even, add.even) # interaction significant, P = 0.002
anova(add.even, site.even) # sponges alone significant, P <.0001
anova(add.even, sponge.even) # site alone not significant, P = 0.1471

# fit variance structure

all.even.var <- update(all.even, weights = varIdent(form = ~ 1 | QuadratSite))

# re-fit final model with REML

even.model <- update(all.even.var, method = "REML")

summary(even.model)

# Phi = 0.4132287

jpeg("//Staff/Home/SCIFAC/rovellal/DocumentsRedir/Projects/Defense/residuals/Ch3_LMM_D_Site_Year_SiteYear.jpeg", width = 11, height = 3.5, units = "in", res = 300) 
par(mfrow = c(1, 3))
plot(resid(even.model, type = "normalized"))
hist(residuals(even.model, type = "normalized"))
qqnorm(residuals(even.model, type = "normalized"))
qqline(residuals(even.model, type = "normalized")) # one outlier
dev.off()

```

The models above seem to have rather good fits. We may attempt to fit an LMM if we think that we must take into account quadrat as a random factor. Be prepared for very different fits though.

## Effects

### Richness

```{r, fig.width = 7.5, fig.height = 2.8}

ef.richness <- data.frame(effect("Sponges * Site", rich.model,  xlevel = 10))

ef.rich.plot <- ggplot()+
  geom_point(data = melt.evenness, aes(x = Sponges, y = Richness, shape = Quadrat), alpha = .3)+
  geom_line(data = ef.richness, aes(x = Sponges, y = fit), color = "royalblue", size = .9)+
  geom_line(data = ef.richness, aes(x = Sponges, y = lower), linetype = "dashed", color = "black", size = .5)+
  geom_line(data = ef.richness, aes(x = Sponges, y = upper), linetype = "dashed", color = "black", size = .5)+
  scale_x_continuous(breaks = seq(0 ,350, 50), limits = c(0, 320))+
  theme_bw()+
  labs(x = expression(paste(Sponge~abundance~(individuals~m^-2))), y = expression(italic(S)))+
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank())+
  facet_grid(~ Site)+
  theme(strip.text.x = element_blank(), strip.background = element_blank())
ef.rich.plot

ggsave("//Staff/Home/SCIFAC/rovellal/DocumentsRedir/Projects/Chapter1/analysis/newpics/richnessEffect.pdf",
       ef.rich.plot, useDingbats = T, width = 8, height = 2.7)
  
```

### Diversity

```{r, fig.width = 7.5, fig.height = 2.8}

ef.diversity <- data.frame(Effect(c("Sponges", "Site"), div.model,  xlevel = 10))

ef.div.plot <- ggplot()+
  geom_point(data = melt.evenness, aes(x = Sponges, y = Diversity, shape = Quadrat), alpha = .3)+
  geom_line(data = ef.diversity, aes(x = Sponges, y = fit), color = "royalblue", size = .9)+
  geom_line(data = ef.diversity, aes(x = Sponges, y = lower), linetype = "dashed", color = "black", size = .5)+
  geom_line(data = ef.diversity, aes(x = Sponges, y = upper), linetype = "dashed", color = "black", size = .5)+
  scale_x_continuous(breaks = seq(0 ,350, 50), limits = c(0, 320))+
  labs(x = expression(paste(Sponge~abundance~(individuals~m^-2))), y = expression(italic(H)))+
  theme_bw()+
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank())+
  facet_grid(~ Site)+
  theme(strip.text.x = element_blank(), strip.background = element_blank())
ef.div.plot

ggsave("//Staff/Home/SCIFAC/rovellal/DocumentsRedir/Projects/Chapter1/analysis/newpics/diversityEffect.pdf",
       ef.div.plot, useDingbats = T, width = 8, height = 2.7)
  
```

### Evenness

```{r, fig.width = 7.5, fig.height = 2.8}

ef.evenness <- data.frame(effect("Sponges * Site", even.model,  xlevel = 10))

ef.even.plot <- ggplot()+
  geom_point(data = melt.evenness, aes(x = Sponges, y = Evenness, shape = Quadrat), alpha = .3)+
  geom_line(data = ef.evenness, aes(x = Sponges, y = fit), color = "royalblue", size = .9)+
  geom_line(data = ef.evenness, aes(x = Sponges, y = lower), linetype = "dashed", color = "black", size = .5)+
  geom_line(data = ef.evenness, aes(x = Sponges, y = upper), linetype = "dashed", color = "black", size = .5)+
  scale_x_continuous(breaks = seq(0 ,350, 50), limits = c(0, 320))+
  scale_y_continuous(breaks = seq(0 ,1, 0.2), limits = c(0, 1))+
  labs(x = expression(paste(Sponge~abundance~(individuals~m^-2))), y = expression(italic(E[1/D])))+
  theme_bw()+
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.minor.x = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank())+
  facet_grid(~ Site)+
  theme(strip.text.x = element_blank(), strip.background = element_blank())
ef.even.plot

ggsave("//Staff/Home/SCIFAC/rovellal/DocumentsRedir/Projects/Chapter1/analysis/newpics/evennessEffect.pdf",
       ef.even.plot, useDingbats = T, width = 8, height = 2.7)
  
```

Intervals above are 95% confidence intervals on the population parameters, i.e. they are not predition intervals. That is why they do not cover the scatter of the points.

## Biodiversity metrics in time.

```{r, fig.width = 7.5, fig.height = 2.8}

ten <- data.frame(rep(NA, 15), rep(10, 15), melt.evenness[melt.evenness$Year == 5, 3:5], 
                  rep(NA, 15), rep(NA, 15),rep(NA, 15))
twelve <- data.frame(rep(NA, 15), rep(12, 15), melt.evenness[melt.evenness$Year == 5, 3:5], 
                     rep(NA, 15), rep(NA, 15),rep(NA, 15))

# rename columns

names(ten) <- colnames(melt.evenness)
names(twelve) <- colnames(melt.evenness)

# bind it all together

evenness.gaps <- rbind.data.frame(melt.evenness[melt.evenness$Year %in% 5:9,],
                                  ten,
                                  melt.evenness[melt.evenness$Year == 11,],
                                  twelve,
                                  melt.evenness[melt.evenness$Year %in% 13:17,])


```

### Richness

```{r, fig.width = 7.5, fig.height = 2.8}

points.richness <- ggplot(data = evenness.gaps, aes(x = Year, y = Richness))+
  geom_line(aes(color = Quadrat), size = .75)+
  geom_point(aes(shape = Quadrat, color = Quadrat), size =2)+
  scale_color_grey(start = 0, end = .7)+
  scale_x_continuous(breaks = seq(5,17,1),
                     labels = seq(2005,2017,1),
                     limits = c(5,17))+
  scale_y_continuous(limits = c(0,35),
                     breaks = seq(0,35,5))+
  labs(x= "Year",
       y = "S")+
  theme_bw()+
  theme(panel.grid.minor = element_blank(), 
        panel.grid.major = element_blank())+
  theme(plot.title = element_text(size=14, vjust=2))+
  theme(axis.title.x = element_text(size=10,vjust=-0.5),
        axis.title.y = element_text(size=10,vjust=0.5))+
  theme(axis.text.x=element_text(size=10, angle = 60, 
                                 hjust = 1, vjust = .9))+
  theme(axis.text.y=element_text(size=10))+
  facet_grid(. ~ Site, scales = "free")+
  theme(strip.text.x = element_blank(), strip.background = element_blank())
points.richness

# ggsave("//Staff/Home/SCIFAC/rovellal/DocumentsRedir/Projects/Chapter 1/pics/richnessDynamics.pdf",
#        points.richness, useDingbats = T, width = 8, height = 3.5)

```

### Evenness

```{r, fig.width = 7.5, fig.height = 2.8}

points.evenness <- ggplot(data = evenness.gaps, aes(x = Year, y = Evenness))+
  geom_line(aes(color = Quadrat), size = .75)+
  geom_point(aes(shape = Quadrat, color = Quadrat), size =2)+
  scale_color_grey(start = 0, end = .7)+
  scale_x_continuous(breaks = seq(5,17,1),
                     labels = seq(2005,2017,1),
                     limits = c(5,17))+
  scale_y_continuous(limits = c(0,1),
                     breaks = seq(0,1,.2))+
  labs(x= "Year",
       y = expression(paste(E[1/D])))+
  theme_bw()+
  theme(panel.grid.minor = element_blank(), 
        panel.grid.major = element_blank())+
  theme(plot.title = element_text(size=14, vjust=2))+
  theme(axis.title.x = element_text(size=10,vjust=-0.5),
        axis.title.y = element_text(size=10,vjust=0.5))+
  theme(axis.text.x=element_text(size=10, angle = 60, 
                                 hjust = 1, vjust = .9))+
  theme(axis.text.y=element_text(size=10))+
  facet_grid(. ~ Site, scales = "free")+
  theme(strip.text.x = element_blank(), strip.background = element_blank())
points.evenness

# ggsave("//Staff/Home/SCIFAC/rovellal/DocumentsRedir/Projects/Chapter 1/pics/evennessDynamics.pdf",
#        points.evenness, useDingbats = T, width = 8, height = 3.5)

```

### Diversity

```{r, fig.width = 7.5, fig.height = 2.8}

points.diversity <- ggplot(data = evenness.gaps, aes(x = Year, y = Diversity))+
  geom_line(aes(color = Quadrat), size = .75)+
  geom_point(aes(shape = Quadrat, color = Quadrat), size =2)+
  scale_color_grey(start = 0, end = .7)+
  scale_x_continuous(breaks = seq(5,17,1),
                     labels = seq(2005,2017,1),
                     limits = c(5,17))+
  scale_y_continuous(limits = c(0,3),
                     breaks = seq(0,3,.5))+
  labs(x= "Year",
       y = expression(paste(E[1/D])))+
  theme_bw()+
  theme(panel.grid.minor = element_blank(), 
        panel.grid.major = element_blank())+
  theme(plot.title = element_text(size=14, vjust=2))+
  theme(axis.title.x = element_text(size=10,vjust=-0.5),
        axis.title.y = element_text(size=10,vjust=0.5))+
  theme(axis.text.x=element_text(size=10, angle = 60, 
                                 hjust = 1, vjust = .9))+
  theme(axis.text.y=element_text(size=10))+
  facet_grid(. ~ Site, scales = "free")+
  theme(strip.text.x = element_blank(), strip.background = element_blank())
points.diversity

# ggsave("//Staff/Home/SCIFAC/rovellal/DocumentsRedir/Projects/Chapter 1/pics/diversityDynamics.pdf",
#        points.diversity, useDingbats = T, width = 8, height = 3.5)

```

